# Write less code, part I

## You know the drill by now

- Many social and data scientists 
    - Code in notebooks like Jupyter or RMarkdown
    - Prioritize code that gets the job done, even if it's wordy
- Many software engineers
    - Code in scripts like .py or .R files
    - Prioritize succinct code, even if a wordier option would be easier to write

I've written a lot of code that technically worked. The data got cleaned, the models got run, and deliverables got sent off on time.

However, until I started taking a more engineering-like approach, I would write *long* chunks of code in notebooks. I know from experience I'm not the only soical or data scientist who leans this way.

## But my code works!

At first this shift might seem pedantic, "why are you bothering me if my code works?"

And while I get where that argument is coming from, two things:

1. The code running without errors ≠ the code working[^1]
2. Writing code for human undestanding > writing code for machine understanding

You don't even need to be working with anyone else on a project to check the human understanding part. How many times have you come back to code you wrote a while ago and struggled to understand what's happening? I know I have. 

Luckily there's a straightforward approach to make your code easier to understand: writing code in smaller chunks.

## But how?

Your code likely needs to do a lot. But if we plan ahead like we discussed in the last chapter we can still break the code down.

Let's revisit our pseudocode from the last chapter

::: {.panel-tabset group="language"}
### R
```{r}
#| eval: false
# Load the data
# Only include penguins that are present on all islands
# Group by species and calculate average bill length
# Arrange by longest bill length and display results
```

### Python
```{python}
#| eval: false
#| python.reticulate: false
# Load the data
# Only include penguins that are present on all islands
# Group by species and calculate average bill length
# Arrange by longest bill length and display results
```
:::

This is a good start, and we can do better. 

One indicator that the code might be doing too much is the presence of conjunctions: and, but, or

Ideally we want each section of code to do one thing at a time. The technical term here is [single responsibility principle](https://en.wikipedia.org/wiki/Single-responsibility_principle).[^2] 

Here's an update to the pseudocode based on trying to do one thing at a time.

::: {.panel-tabset group="language"}
### R
```{r}
#| eval: false
# Load the data
# Only include penguins that are present on all islands
# Group by species
# Calculate average bill length
# Arrange by longest bill length
# Display results
```

### Python
```{python}
#| eval: false
#| python.reticulate: false
# Load the data
# Only include penguins that are present on all islands
# Group by species
# Calculate average bill length
# Arrange by longest bill length
# Display results
```
:::

Now our pseudocode sets us up to write the code in smaller chunks. I suggest literally creating separate chunks for each line of pseudocode code if you're writing code in a notebook.

## Why bother?

This breakout might seem excessive at first, but we're spending a bit more time up front to make our code far easier to understand/debug later. 

To take an extreme example, switch the code to your less familiar language. This version probably feels daunting to read:

::: {.panel-tabset group="language"}
### R
```{r}
#| eval: false
library(palmerpenguins)
library(dplyr)

penguins |> filter(species %in% (penguins |> group_by(species) |> summarize(island_count = n_distinct(island)) |> filter(island_count == max((penguins |> summarize(max_islands = n_distinct(island)))$max_islands)) |> pull(species))) |> group_by(species) |> summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE)) |> arrange(desc(avg_bill_length)) |> print()
```

### Python
```{python}
#| eval: false
#| python.reticulate: false
import pandas as pd
from palmerpenguins import load_penguins

penguins = load_penguins()
print(penguins.groupby('species').filter(lambda x: x['species'].iloc[0] in penguins.groupby('species')['island'].nunique()[penguins.groupby('species')['island'].nunique() == penguins['island'].nunique()].index).groupby('species')['bill_length_mm'].mean().sort_values(ascending=False))
```
:::

While this version can feel more straightforward:

::: {.panel-tabset group="language"}
### R
```{r}
#| eval: false
# Load the data
library(palmerpenguins)
library(dplyr)
penguins_data <- penguins

# Only include penguins that are present on all islands
species_island_counts <- penguins_data |>
  group_by(species) |>
  summarize(island_count = n_distinct(island))

total_islands <- n_distinct(penguins_data$island)

species_on_all_islands <- species_island_counts |>
  filter(island_count == total_islands) |>
  pull(species)

penguins_filtered <- penguins_data |>
  filter(species %in% species_on_all_islands)

# Group by species
penguins_grouped <- penguins_filtered |>
  group_by(species)

# Calculate average bill length
penguins_with_avg <- penguins_grouped |>
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))

# Arrange by longest bill length
penguins_sorted <- penguins_with_avg |>
  arrange(desc(avg_bill_length))

# Display results
print(penguins_sorted)
```

### Python
```{python}
#| eval: false
#| python.reticulate: false
# Load the data
import pandas as pd
from palmerpenguins import load_penguins

penguins_data = load_penguins()

# Only include penguins that are present on all islands
species_island_counts = penguins_data.groupby('species')['island'].nunique()
total_islands = penguins_data['island'].nunique()
species_on_all_islands = species_island_counts[species_island_counts == total_islands].index
penguins_filtered = penguins_data[penguins_data['species'].isin(species_on_all_islands)]

# Group by species
penguins_grouped = penguins_filtered.groupby('species')

# Calculate average bill length
penguins_with_avg = penguins_grouped['bill_length_mm'].mean().reset_index()
penguins_with_avg.columns = ['species', 'avg_bill_length']

# Arrange by longest bill length
penguins_sorted = penguins_with_avg.sort_values('avg_bill_length', ascending=False)

# Display results
print(penguins_sorted)
```
:::

## Is this actually less code?

Even if we just stopped here, our code would still be easier to read + debug later.

But to truly write less code, we need to learn two new concepts we'll introduce in the next chapter.

For example, right now we're relying on comments to tell us what we're doing. That's not inherently bad, and generally we want to save comments to explain *why* we're doing something.

Also, this granular approach feels like a lot if we have to do it more than once. What if we also want to look at flipper length in addition to bill length? And right now we're writing *more* lines of code vs. the less readable option.

## Practice breaking down code into chunks

::: {.blendtutor language="python" packages="pandas"}
The one-liner in the template works, but it does everything at once. Rewrite it
so each chunk does one thing, following the pseudocode comments. Store the final
table in `penguins_sorted` with columns `species` and `avg_bill_length`.

```python
import pandas as pd

penguins_data = pd.DataFrame({
    "species": ["Adelie"] * 3 + ["Gentoo"] * 3 + ["Chinstrap"] * 2,
    "island": ["Biscoe", "Dream", "Torgersen"] * 2 + ["Dream"] * 2,
    "bill_length_mm": [38.0, 40.0, 39.0, 47.0, 49.0, 48.0, 50.0, 52.0],
})

# Works, but hard to read:
# print(penguins_data.groupby('species').filter(lambda x: x['species'].iloc[0] in penguins_data.groupby('species')['island'].nunique()[penguins_data.groupby('species')['island'].nunique() == penguins_data['island'].nunique()].index).groupby('species')['bill_length_mm'].mean().sort_values(ascending=False))

# Only include penguins that are present on all islands

# Group by species

# Calculate average bill length

# Arrange by longest bill length

# Display results
```

```{.python .checks}
assert list(penguins_sorted['species']) == ['Gentoo', 'Adelie']
assert list(penguins_sorted['avg_bill_length']) == [48.0, 39.0]
```

```{.python .solution}
import pandas as pd

penguins_data = pd.DataFrame({
    "species": ["Adelie"] * 3 + ["Gentoo"] * 3 + ["Chinstrap"] * 2,
    "island": ["Biscoe", "Dream", "Torgersen"] * 2 + ["Dream"] * 2,
    "bill_length_mm": [38.0, 40.0, 39.0, 47.0, 49.0, 48.0, 50.0, 52.0],
})

# Only include penguins that are present on all islands
species_island_counts = penguins_data.groupby("species")["island"].nunique()
total_islands = penguins_data["island"].nunique()
species_on_all_islands = species_island_counts[species_island_counts == total_islands].index
penguins_filtered = penguins_data[penguins_data["species"].isin(species_on_all_islands)]

# Group by species
penguins_grouped = penguins_filtered.groupby("species")

# Calculate average bill length
penguins_with_avg = penguins_grouped["bill_length_mm"].mean().reset_index()
penguins_with_avg.columns = ["species", "avg_bill_length"]

# Arrange by longest bill length
penguins_sorted = penguins_with_avg.sort_values("avg_bill_length", ascending=False)

# Display results
print(penguins_sorted)
```

::: {.hints}
- Give each step its own named variable so you can print and check it on its own.
- Count islands per species with `.groupby("species")["island"].nunique()`.
- `.isin()` keeps the rows whose species is in a list of names.
:::

::: {.success-criteria}
- Each chunk does one thing (no step joined by "and"), matching the pseudocode
- Intermediate results are stored in named variables, not chained in one line
- `penguins_sorted` holds only species present on all islands, with their average bill length, longest first
:::
:::

[^1]: More on this idea in the "Write more code" chapters
[^2]: The Wiki here leans toward something called object-oriented programming, but this idea extends beyond that.

